Mountaineering: Be lucky first, prepared second

### 2024-12-14 paper/paper.qmd
update clean data

---
title: "Mountaineering: Be lucky first, prepared second"
subtitle: "Preperation and Timing are more important than Gender Identity, Age, Amount of expedition members, and height of peak when mountaineering in the Himalayas."
author: 
  - Daniel Xu
thanks: "Code and data are available at: [https://github.com/possibleburger2/](https://github.com/possibleburger2/Himalayan-Climbing-Expeditions-Data-Analysis)."
date: today
date-format: long
abstract: "This study analyzes over a century of Himalayan mountaineering expeditions (1905–2019) to identify key factors influencing success. Using logistic regression, we find that preparation and timing—such as the use of supplemental oxygen and climbing during optimal seasons—are more important predictors of success than individual traits like age, gender, or the height of the peak. These findings reveal that success in high-altitude climbing depends more on strategic choices than personal attributes. This insight underscores the value of informed planning and environmental awareness for achieving mountaineering goals."
format: pdf
number-sections: true
bibliography: references.bib

---

@@

## Overview
The raw datasets were obtained from @Cookson, who sourced them from The Himalayan Database [@Salisbury].

@@

@@

peaks |>
  ggplot(aes(climbing_status, fill = climbing_status)) +
  geom_bar() +
  labs(
    title = "More than a quarter of Himalayan peaks remain unclimbed",
    x = "",
    y = "Number of peaks"
  ) +
  theme(legend.position = "none")
```
This graph shows that despite over 100 years of climbing history, not all attempted peaks have been conquered.
Why do so many climbs fail however? The following graph shows the most common reasons.
\newpage

@@

@tbl-termination shows that most attempts fail due to partially random events like weather, or random events such Avalanches. 
\newpage
```{r, fig.width = 12, fig.height = 3}
#| warning: false
#| message: false
#| echo: false
#| label: fig-summit
#| fig-cap: "Distribution of First ascents across years"

@@

countries_by_decade |>
  ggplot(aes(first_ascent_decade, pct_of_ascents, fill = first_ascent_country)) +
  geom_col() +
  scale_x_continuous(breaks = seq(1930, 2010, 20)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +  # Correct label function
  facet_wrap(~ first_ascent_country) +
  labs(
    title = "Nepal has been consistently involved in first ascents",
    subtitle = "Percent of first ascents involving a countries' citizens",
    x = "Decade of first ascent",
    y = ""
  ) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(colour = "black"),
    strip.background = element_blank()
  )
```
Notably there was only one data point past 1930, where X expedition had an ascent for X mountain in 1905 that was left out to save space from the UK so there should be a single bar only for the UK at 100% at 1905.

@@

## Measurement
The measurement of Himalayan climbing data in the referenced dataset involves converting real-world phenomena—such as expeditions, summits, and fatalities—into structured entries. This process begins with records collected from expedition reports, which are typically submitted by climbers or organizations. These reports include details about expedition dates, routes, climbers' nationalities, success in reaching the summit, and any fatalities or injuries that occurred.

@@

## Model {#sec-model}
I used a Binary Logistic Regression uses the age of an individual,the amount of members in their expedition, the amount of oxygen they used, the season of their climb, their sex and the height of mountain to determine the likelihood of an individual of summit-ting the peak or "succeeding". A Binary Logistic Regression is a statistical method to predict binary outcomes like the success or failure of a climb based on both categorical and continuous variables.
The Model uses six independent demographic variables: "age","members","oxygen_used","season","sex", and "height_metres" and the dependent variable is "success"
The logistic regression model I will be using is:
$$
\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{age} + \beta_2 \times \text{members} + \beta_3 \times \text{oxygenused} + \beta_4 \times \text{season} + \beta_5 \times \text{sex} + \beta_6 \times \text{heightmetres} 
$$

## Model Justification
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary1

@@

#modelsummary(model2,statistic = c("p.value")) simply took output and inputted as png, the length of the table is too long and won't wrap
```
\newpage

@@

Because we have so many statistically insignificant variables at a 5% uncertainty we should try to remove them to see the impact it has on the model
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary2

modelsummary(model1, statistic = c("p.value"))
```
\newpage
The second table shows the results when citizenship and hired_staff are removed. We see the RMSE only grains 0.01, meaning it loses 1% accuracy. This means the citizenship of climber nor the amount of hired staff did not contribute to the model's accuracy and were good to remove.
```{r, fig.width = 2, fig.height = 1}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary

results <- anova(model1,model2,
      test = "LRT")
kable(results, caption = "ANOVA Model Comparison") |>
  kable_styling(full_width = FALSE, position = "center")
```
This anova table just confirms that there is no difference in the modelling ability after removing the two predictor variables.

## Results {#sec-results}
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-model-coefficients
#| fig-cap: Coefficients of the model

@@

# Plot for the coefficients of the model
ggplot(coefficients, aes(x = estimate, y = term)) +
  geom_point() + # Coefficient points
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) + # Confidence/Credible interval bars
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + # Add zero line
  labs(x = "Coefficient Estimate", y = "Predictor Variable") + # Axis labels
  theme_minimal() # Minimal theme for better readability
``` 
@fig-model-coefficients shows most of the coefficients do not have a large affect on success however have a very certain affect on success. The season you go are all relative to fall, so going during the fall has a large measured impact on success as well as oxygen used and sex.

@@

## Chances of success decline with increase in age
Age is a significant factor in mountaineering success, with success rates generally decreasing as climbers get older. This decline can be attributed to various age-related factors, such as the natural loss of muscle mass, strength, and aerobic capacity. As climbers age, they may also experience reduced agility, flexibility, and slower recovery, which can hinder their performance on difficult terrain and in recovering from intense physical exertion.

@@

## Weaknesses and Limitations
While my analysis offers valuable insights into the factors influencing a successful summit, it is important to acknowledge the weaknesses and limitations.

@@

# Appendix {#sec-appendix}
## Cleaning
For the analysis data, the cleaning steps were from @Cookson

@@

## Observational data
Himalayan climbing data, particularly concerning summits and fatalities on peaks like Mount Everest, may misrepresent actual outcomes due to several factors, including incomplete reporting, commercial biases, and evolving definitions of success. These discrepancies can create a skewed understanding of the risks and achievements in high-altitude mountaineering.
Incomplete Reporting

>>>>

---
title: "Mountaineering: Be lucky first, prepared second"
subtitle: "Preperation and Timing are more important than Gender Identity, Age, Amount of expedition members, and height of peak when mountaineering in the Himalayas."
author: 
  - Daniel Xu
thanks: "Code and data are available at: [https://github.com/possibleburger2/](https://github.com/possibleburger2/Himalayan-Climbing-Expeditions-Data-Analysis)."
date: today
date-format: long
abstract: "This study analyzes over a century of Himalayan mountaineering expeditions (1905–2019) to identify key factors influencing success. Using logistic regression, we find that preparation and timing—such as the use of supplemental oxygen and climbing during optimal seasons—are more important predictors of success than individual traits like age, gender, or the height of the peak. These findings reveal that success in high-altitude climbing depends more on strategic choices than personal attributes. This insight underscores the value of informed planning and environmental awareness for achieving mountaineering goals."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: sentence
---

@@

## Overview

The raw datasets were obtained from @Cookson, who sourced them from The Himalayan Database [@Salisbury].

@@

Members of expedition

```{r}

#| warning: false
#| message: false
#| echo: false
#| label: fig-members
#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."

expeditions |>
  ggplot(aes(members)) +
  geom_histogram(
    binwidth = 2,
    alpha = 0.8
  ) +
  labs(
    title = "Distribution of Expedition members",
    caption = "Source: The Himalayan Database",
    x = "Height (m) above sea level",
    y = "Number of peaks"
  )
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-members
#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."

expeditions$success <- expeditions$termination_reason %in% c("Success (main peak)", "Success (subpeak)")
success_rate <- expeditions |>
  group_by(members)|>
  summarize(success_rate = mean(success))
ggplot(success_rate, aes(x = members, y = success_rate)) +
  geom_line(color = "blue") + # Line plot
  geom_point(color = "red") + # Points to emphasize values
  labs(
    title = "Success Rate by Members",
    x = "Number of Members",
    y = "Success Rate"
  ) +
  theme_minimal()
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-season
#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."

ggplot(expeditions,aes(season)) +
  geom_bar()
```

@@

peaks |>
  ggplot(aes(climbing_status, fill = climbing_status)) +
  geom_bar() +
  labs(
    title = "More than a quarter of Himalayan peaks remain unclimbed",
    x = "",
    y = "Number of peaks"
  ) +
  theme(legend.position = "none")
```

This graph shows that despite over 100 years of climbing history, not all attempted peaks have been conquered.
Why do so many climbs fail however?
The following graph shows the most common reasons.
\newpage

@@

@tbl-termination shows that most attempts fail due to partially random events like weather, or random events such Avalanches.
\newpage

```{r, fig.width = 12, fig.height = 3}
#| warning: false
#| message: false
#| echo: false
#| label: fig-summit
#| fig-cap: "Distribution of First ascents across years"

@@

countries_by_decade |>
  ggplot(aes(first_ascent_decade, pct_of_ascents, fill = first_ascent_country)) +
  geom_col() +
  scale_x_continuous(breaks = seq(1930, 2010, 20)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +  # Correct label function
  facet_wrap(~ first_ascent_country) +
  labs(
    title = "Nepal has been consistently involved in first ascents",
    subtitle = "Percent of first ascents involving a countries' citizens",
    x = "Decade of first ascent",
    y = ""
  ) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(colour = "black"),
    strip.background = element_blank()
  )
```

Notably there was only one data point past 1930, where X expedition had an ascent for X mountain in 1905 that was left out to save space from the UK so there should be a single bar only for the UK at 100% at 1905.

@@

## Measurement

The measurement of Himalayan climbing data in the referenced dataset involves converting real-world phenomena—such as expeditions, summits, and fatalities—into structured entries.
This process begins with records collected from expedition reports, which are typically submitted by climbers or organizations.
These reports include details about expedition dates, routes, climbers' nationalities, success in reaching the summit, and any fatalities or injuries that occurred.

@@

## Model {#sec-model}

I used a Binary Logistic Regression uses the age of an individual,the amount of members in their expedition, the amount of oxygen they used, the season of their climb, their sex and the height of mountain to determine the likelihood of an individual of summit-ting the peak or "succeeding".
A Binary Logistic Regression is a statistical method to predict binary outcomes like the success or failure of a climb based on both categorical and continuous variables.
The Model uses six independent demographic variables: "age","members","oxygen_used","season","sex", and "height_metres" and the dependent variable is "success" The logistic regression model I will be using is: $$
\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{age} + \beta_2 \times \text{members} + \beta_3 \times \text{oxygenused} + \beta_4 \times \text{season} + \beta_5 \times \text{sex} + \beta_6 \times \text{heightmetres} 
$$

## Model Justification

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary1

@@

#modelsummary(model2,statistic = c("p.value")) simply took output and inputted as png, the length of the table is too long and won't wrap
```

@@

Because we have so many statistically insignificant variables at a 5% uncertainty we should try to remove them to see the impact it has on the model

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary2

modelsummary(model1, statistic = c("p.value"))
```

The second table shows the results when citizenship and hired_staff are removed.
We see the RMSE only grains 0.01, meaning it loses 1% accuracy.
This means the citizenship of climber nor the amount of hired staff did not contribute to the model's accuracy and were good to remove.

```{r, fig.width = 2, fig.height = 1}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary

results <- anova(model1,model2,
      test = "LRT")
kable(results, caption = "ANOVA Model Comparison") |>
  kable_styling(full_width = FALSE, position = "center")
```

This anova table just confirms that there is no difference in the modelling ability after removing the two predictor variables.

## Results {#sec-results}

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-model-coefficients
#| fig-cap: Coefficients of the model

@@

# Plot for the coefficients of the model
ggplot(coefficients, aes(x = estimate, y = term)) +
  geom_point() + # Coefficient points
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) + # Confidence/Credible interval bars
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + # Add zero line
  labs(x = "Coefficient Estimate", y = "Predictor Variable") + # Axis labels
  theme_minimal() # Minimal theme for better readability
```

@fig-model-coefficients shows most of the coefficients do not have a large affect on success however have a very certain affect on success.
The season you go are all relative to fall, so going during the fall has a large measured impact on success as well as oxygen used and sex.

@@

## Chances of success decline with increase in age

Age is a significant factor in mountaineering success, with success rates generally decreasing as climbers get older.
This decline can be attributed to various age-related factors, such as the natural loss of muscle mass, strength, and aerobic capacity.
As climbers age, they may also experience reduced agility, flexibility, and slower recovery, which can hinder their performance on difficult terrain and in recovering from intense physical exertion.

@@

## Weaknesses and Limitations

While my analysis offers valuable insights into the factors influencing a successful summit, it is important to acknowledge the weaknesses and limitations.

@@

# Appendix {#sec-appendix}

## Cleaning

For the analysis data, the cleaning steps were from @Cookson

@@

## Observational data

Himalayan climbing data, particularly concerning summits and fatalities on peaks like Mount Everest, may misrepresent actual outcomes due to several factors, including incomplete reporting, commercial biases, and evolving definitions of success.
These discrepancies can create a skewed understanding of the risks and achievements in high-altitude mountaineering.
Incomplete Reporting
